Legacy Data Confounds Modern Genomics Studies

### 2019-06-12 LatexManuscript/legacy.tex
final update
many updates for reviewer comments

To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
Traditionally, genome wide association studies use genotypes as the independent variable. 
Here we perform a ``reverse GWAS'', in the sense that genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}A).
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.

@@

\begin{table}[h!]
\centering
\begin{tabular}{l  r r r}
                      & {Repeat}  & {Non-Repeat}   & {Total}    \\ \hline
{SNP}  & 3,369 & 11,059 & 14,428\\  
{INDEL} & 181 & 657 & 838\\ \hline
{Total} & 3,550 & 11,716 & 15,270\\
\end{tabular}
\caption{Number of statistically significant variants  passing the 1000 Genomes Project strict mask per category. 
Variants that are flagged by the 1000 Genomes Project nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
A total of 15,270 are statistically significantly associated to $Q$. 
The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}
\label{sigTable}
\end{table}

@@

In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}. To assess consistency between the two datasets, we consider consistency of genotype calls for $Q$-associated variants that are predicted to be polymorphic in these 83 individuals according to the 1kGP.
Among the 296 such variants that were $Q$-associated in the CHB or CHS, only 6 are present in the resequenced data (Figure \ref{90HanSFS}). 
This is more than our nominal false positive rate of 1\% of the sites. Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes.

We did a similar analysis using all variants identified in the GCAT model (rather than only variants significantly associated to $Q$ within the CHB and CHS). 
Of the 15,270 $Q$-associated variants identified globally, 6,307 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
From this subset, only 1,139 (or 18\%) are present in the resequenced data.
The allele frequencies of these variants are nearly identical between datasets suggesting that among these 83 individuals, these variants are properly genotyped in the 1kGP.
There are 5 alleles that show differing frequencies between both datasets that are likely explained by biased genotypes.
The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, supporting sequencing rather than cell line artifacts.

@@

\begin{table}[h!]
\begin{tabular}{l l l r r r}
  {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$\\-$\log_{10} p$\\(adjusted)}      \\ \hline
28654678 & PLoS One & rs201761909 & 5.7 & 78.11\\
	& 	&  	rs201130852 & 5.05 & 72.28\\
	& 	&  	rs201255786 & 5.7 & 68.97\\
	& 	&  	rs200655768 & 6.52 & 66.67\\
	& 	&  	rs184202621 & 5.52 & 60.45\\
	& 	& 	rs80274284 & 6 & 56.15\\
	& 	& 	rs200699422 & 5.3 & 7.43\\
23527680 & \multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics} & rs6057648 & 5.4 & 20.5\\
28928442 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & rs201471471 & 6.52 & 7.87\\
26053186 & \multicolumn{1}{p{3cm}}{\raggedright PLoS One} & rs60136336 & 5.7 & 2.25\\ \hline
\textcolor{gray}{28270201} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Genome Medicine}} & \textcolor{gray}{rs453755} & \textcolor{gray}{7.52} & \textcolor{gray}{5.29}\\
\textcolor{gray}{23023329} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*15.3} & \textcolor{gray}{4.32}\\
\textcolor{gray}{28334899} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Human Molecular Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*29.3} & \textcolor{gray}{4.32}\\
\textcolor{gray}{28240269} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Communications}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*72.7} & \textcolor{gray}{4.32}\\
\textcolor{gray}{27863252} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Cell}\\} & \textcolor{gray}{rs3794738} & \textcolor{gray}{*13.15} & \textcolor{gray}{3.73}\\
\textcolor{gray}{29534301} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Hepatology}\\} & \textcolor{gray}{rs9273062} & \textcolor{gray}{*9.7} & \textcolor{gray}{3.36}\\
\textcolor{gray}{21386085} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Diabetes}\\}  & \textcolor{gray}{rs301} & \textcolor{gray}{*10.52} & \textcolor{gray}{3.02}\\
\textcolor{gray}{26830138} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Molecular Psychiatry}} & \textcolor{gray}{rs77894924} & \textcolor{gray}{6.7} & \textcolor{gray}{2.77}\\
\textcolor{gray}{29617998} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Human Molecular Genetics}} & \textcolor{gray}{rs4963156} & \textcolor{gray}{*22.4} & \textcolor{gray}{2.52}\\
\textcolor{gray}{28698626} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Scientific Reports}} & \textcolor{gray}{rs11015915} & \textcolor{gray}{5.05} & \textcolor{gray}{2.45}\\
\textcolor{gray}{26974007} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs3124998} & \textcolor{gray}{*8.05} & \textcolor{gray}{2.33}\\
\textcolor{gray}{26634245} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{BMC Genetics}} & \textcolor{gray}{rs451000} & \textcolor{gray}{6} & \textcolor{gray}{2.28}\\
 	& 	& 	\textcolor{gray}{rs443874} & \textcolor{gray}{5.3} & \textcolor{gray}{2.26}\\
	& 	&	\textcolor{gray}{rs400942} & \textcolor{gray}{6} & \textcolor{gray}{2.2}\\
\textcolor{gray}{25918132} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Toxicological Sciences}} & \textcolor{gray}{rs76780579} & \textcolor{gray}{6} & \textcolor{gray}{2.09}\\

@@

On a technical front, we were surprised that strong association between variants and technical covariates in the 1kGP project had not been identified before. 
The genome-wide logistic regression analysis of genotype on quality metric is straightforward, and should probably be a standard in a variety of -omics studies. The logistic factor analysis is more computationally demanding but produces more robust results \sgcomment{We never explained this!}. Both approaches produce comparable results.

>>>>

To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
Traditionally, genome wide association studies use genotypes as the independent variable. 
Here we perform a \luke{genotype conditional association test (GCAT), where} genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}A).
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.

@@

\begin{table}[h!]
\centering
\begin{tabular}{l  r r r}
                      & {Repeat}  & {Non-Repeat}   & {Total}    \\ \hline
{SNP}  & 3,369 \luke{(0.53\textperthousand)} & 11,059 (0.56\textperthousand) & 14,428\\  
{INDEL} & 181 (0.3\textperthousand) & 657 (0.66\textperthousand) & 838\\ \hline
{Total} & 3,550 & 11,716 & 15,270\\
\end{tabular}
\caption{Number of statistically significant variants  passing the 1000 Genomes Project strict mask per category. 
Variants that are flagged by the 1000 Genomes Project nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
A total of 15,270 are statistically significantly associated to $Q$. 
The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}
\label{sigTable}
\end{table}

@@

In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}. To assess consistency between the two datasets, we consider consistency of genotype calls for $Q$-associated variants that are predicted to be polymorphic in these 83 individuals according to the 1kGP.
Among the 296 such variants that were $Q$-associated in the CHB or CHS, only 6 are present in the resequenced data (Figure \ref{90HanSFS}). 
This is more than our nominal false discovery rate of \luke{$\alpha = 0.01$} of the sites. 
Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes.

We did a similar analysis using all variants identified in the GCAT model (rather than only variants significantly associated to $Q$ within the CHB and CHS). 
Of the 15,270 $Q$-associated variants identified globally, 6,307 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
From this subset, only 1,139 (or 18\%) are present in the resequenced data.
The allele frequencies of these variants are nearly identical between datasets suggesting that among these 83 individuals, these variants are properly genotyped in the 1kGP.
There are 5 \luke{variants} that show differing frequencies between both datasets that are likely explained by biased genotypes.
The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, supporting sequencing rather than cell line artifacts.

@@

\begin{table}[h!]
\begin{tabular}{l l l r r r}
  {Pubmed ID}  & {Disease/Trait} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$\\-$\log_{10} p$\\(adjusted)}      \\ \hline
28654678 & EBV copy number in & rs201761909 & 5.7 & 78.11\\
	&   lymphoblastoid cell lines	&  	rs201130852 & 5.05 & 72.28\\
	& 	&  	rs201255786 & 5.7 & 68.97\\
	& 	&  	rs200655768 & 6.52 & 66.67\\
	& 	&  	rs184202621 & 5.52 & 60.45\\
	& 	& 	rs80274284 & 6 & 56.15\\
	& 	& 	rs200699422 & 5.3 & 7.43\\
23527680 & \multicolumn{1}{p{3cm}}{\raggedright Attention deficit\\hyperactivity disorder} & rs6057648 & 5.4 & 20.5\\
28928442 & \multicolumn{1}{p{3cm}}{\raggedright Cold sores} & rs201471471 & 6.52 & 7.87\\
26053186 & \multicolumn{1}{p{3cm}}{\raggedright HMPMA levels in smokers} & rs60136336 & 5.7 & 2.25\\ \hline
\textcolor{gray}{28270201} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{HDL cholesterol}} & \textcolor{gray}{rs453755} & \textcolor{gray}{7.52} & \textcolor{gray}{5.29}\\
\textcolor{gray}{23023329} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Prostate cancer}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*15.3} & \textcolor{gray}{4.32}\\
\textcolor{gray}{28334899} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{HDL cholesterol}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*29.3} & \textcolor{gray}{4.32}\\
\textcolor{gray}{28240269} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Blood protein levels}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*72.7} & \textcolor{gray}{4.32}\\
\textcolor{gray}{27863252} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{High light scatter\\reticulocyte count}\\} & \textcolor{gray}{rs3794738} & \textcolor{gray}{*13.15} & \textcolor{gray}{3.73}\\
\textcolor{gray}{29534301} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Response to\\hepatitis B vaccine}\\} & \textcolor{gray}{rs9273062} & \textcolor{gray}{*9.7} & \textcolor{gray}{3.36}\\
\textcolor{gray}{21386085} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Metabolic syndrome}\\}  & \textcolor{gray}{rs301} & \textcolor{gray}{*10.52} & \textcolor{gray}{3.02}\\
\textcolor{gray}{26830138} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Alzheimer disease\\and age of onset}} & \textcolor{gray}{rs77894924} & \textcolor{gray}{6.7} & \textcolor{gray}{2.77}\\
\textcolor{gray}{29617998} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Intraocular pressure}} & \textcolor{gray}{rs4963156} & \textcolor{gray}{*22.4} & \textcolor{gray}{2.52}\\
\textcolor{gray}{28698626} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Immunoglobulin A vasculitis}} & \textcolor{gray}{rs11015915} & \textcolor{gray}{5.05} & \textcolor{gray}{2.45}\\
\textcolor{gray}{26974007} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Chronic inflammatory diseases}} & \textcolor{gray}{rs3124998} & \textcolor{gray}{*8.05} & \textcolor{gray}{2.33}\\
\textcolor{gray}{26634245} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Post bronchodilator FEV1/FVC ratio}} & \textcolor{gray}{rs451000} & \textcolor{gray}{6} & \textcolor{gray}{2.28}\\
 	& 	& 	\textcolor{gray}{rs443874} & \textcolor{gray}{5.3} & \textcolor{gray}{2.26}\\
	& 	&	\textcolor{gray}{rs400942} & \textcolor{gray}{6} & \textcolor{gray}{2.2}\\
\textcolor{gray}{25918132} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Diisocyanate-induced asthma}} & \textcolor{gray}{rs76780579} & \textcolor{gray}{6} & \textcolor{gray}{2.09}\\

@@

On a technical front, we were surprised that strong association between variants and technical covariates in the 1kGP project had not been identified before. 
The genome-wide logistic regression analysis of genotype on quality metric is straightforward, and should probably be a standard in a variety of -omics studies. The logistic factor analysis is more computationally demanding but produces more robust results \citep{song2015testing}. Both approaches produce comparable results.

### 2019-06-12 PlottingScripts/ProcessPlotData.Rmd
many updates for reviewer comments

#get reverse complement of context, ref and alt
data$Rev<-ReverseComp(as.character(data$Context))
data$RevRef<-ReverseComp(as.character(data$Ref))
data$RevAlt<-ReverseComp(as.character(data$Alt))
#only get one half (to fold over)
AC<-data[which(data$Ref 
               %in% 
                 c('A','C')),]
TG<-data[which(data$Ref 
               %in% 
                 c('T','G')),]
#fold over
TG$Context <- TG$Rev
TG$Ref <- TG$RevRef
TG$Alt <- TG$RevAlt
#bind them together
data <- rbind(AC,TG)
#combine it all together to get a context
data$Mut <- paste0(data$Ref,'->', 
                substr(data$Context,1,1),
                data$Alt,
                substr(data$Context,3,3))
#get the counts of each bin
binCount <- as.data.frame(table(data$Mut))
#process for plotting
binCount$Start <- substr(binCount$Var1,4,4)
binCount$End <- substr(binCount$Var1,6,6)
binCount$Mut <- paste(substr(binCount$Var1,1,3),
                      substr(binCount$Var1,5,5))
return(binCount)
}

@@

>>>>

#get reverse complement of context, ref and alt
data$Rev<-ReverseComp(as.character(data$Context))
data$RevRef<-ReverseComp(as.character(data$Ref))
data$RevAlt<-ReverseComp(as.character(data$Alt))
#only get one half (to fold over)
AC<-data[which(data$Ref 
               %in% 
                 c('A','C')),]
TG<-data[which(data$Ref 
               %in% 
                 c('T','G')),]
#fold over
TG$Context <- TG$Rev
TG$Ref <- TG$RevRef
TG$Alt <- TG$RevAlt
#bind them together
data <- rbind(AC,TG)
#combine it all together to get a context
data$Mut <- paste0(data$Ref,'->', 
                substr(data$Context,1,1),
                data$Alt,
                substr(data$Context,3,3))
#get the counts of each bin
binCount <- as.data.frame(table(data$Mut))
#process for plotting
binCount$Start <- substr(binCount$Var1,4,4)
binCount$End <- substr(binCount$Var1,6,6)
binCount$Mut <- paste(substr(binCount$Var1,1,3),
                      substr(binCount$Var1,5,5))
binCount$Mut <- gsub('->', ' →',binCount$Mut)

binCount$Var1 <- NULL

return(binCount)
}

@@

#Chi-square test of independence
chi_test <- merge(SNP, rSNP, 
                  by = c('Start','End', 'Mut'), 
                  all =T)

chi_test$chi <- 
  (chi_test$Freq.x - chi_test$Freq.y)^2 / chi_test$Freq.y

chi_test$p_val <- pchisq(chi_test$chi, 
                         df=96, 
                         lower.tail=FALSE)

SNP <- merge(SNP, 
             chi_test %>% 
               select(c('Start','End', 'Mut',
                        'chi', 'p_val')), 
             by = c('Start','End', 'Mut'))
